\documentclass{amsart}

\usepackage{jon-thm}
\usepackage{fourier}
\usepackage{tgschola}
\usepackage{xparse}
\usepackage{jon-tikz}

\DeclareMathOperator\colim{colim}
% \DeclareMathOperator\lim{lim}

\title{The theory of logoi}
\author{Mathieu Anel (notes by Jon Sterling)}
\begin{document}
\maketitle
\tableofcontents

In these notes, everything is implicitly $\infty$-categorical; 1-categories
are explicitly marked.

\section{The topo-logie duality}
A topos is a space defined by a \emph{logos of sheaves}; a logos is an algebraic structure on the category of sheaves on a space.


\section{Settling size issues}
The dichotomy between sets and classes is not enough to handle category theory.


\NewDocumentCommand\SET{}{\mathbf{Set}}

\begin{definition}

  An inaccessible cardinal is a cardinal $\alpha$ such that $\SET_{<\alpha}$ is
  stable under dependent product and sum. We will use the terms atomic, finite
  (cat), small (Cat), normal (CAT), large for strictly increasing scales.

\end{definition}

\begin{example}
  The category of small categories is a normal category.
\end{example}


\section{Cocompletion}
\NewDocumentCommand\CC{}{\mathbb{C}}
\NewDocumentCommand\Psh{m}{\mathbf{P}(#1)}
\NewDocumentCommand\BigPsh{m}{\overline{\mathbf{P}}(#1)}
\NewDocumentCommand\Sp{}{\mathcal{S}}
\NewDocumentCommand\Hom{mm}{[#1,#2]}
\NewDocumentCommand\OpCat{m}{#1^{\mathrm{op}}}
\NewDocumentCommand\cat{}{\mathbf{cat}}
\NewDocumentCommand\Cat{}{\mathbf{Cat}}
\NewDocumentCommand\CAT{}{\mathbf{CAT}}
\NewDocumentCommand\CAAT{}{\mathbb{CAT}}
\NewDocumentCommand\Pres{}{\mathbf{Pres}}
\NewDocumentCommand\Loc{mm}{\mathbf{L}(#1,#2)}

\begin{definition}
  A category $C$ is \emph{cocomplete} if for any small category $K$, the
  functor of constant diagrams $\Mor{C}{C^K}$ has a left adjoint.
\end{definition}

A functor $\Mor{C}{D}$ between cocomplete functors should be
\emph{cocontinuous}; this gives a category of cocomplete categories.
%
The forgetful functor from cocomplete categories to complete categories has a
left adjoint (Lurie). This is the \emph{free cocompletion}.

If $C$ is small, the $\Psh{C} = \Hom{\OpCat{C}}{\Sp}$ is a category. If $C$ is
normal, then we take presheaves $\BigPsh{C}$ into $\overline{\Sp}$ and take a subcategory;
we take the smallest subcategory of this big presheaf category which is closed
under small colimits.

\begin{definition}
  $C$ is \emph{presentable} if there exists a small diagram $\Mor{K}{C}$ and a
  small diagram $W\to\Psh{K}^{\to}$, such that $\Loc{\Psh{K}}{W}\simeq C$; in
  other words, $C$ is the localization of a presheaf category at some class of
  maps.
\end{definition}

\begin{definition}
  If $C:\CAT_\alpha$, $X:C$ is \emph{small} if for all filtered
  $\Mor[Y]{K}{C}$, we have $\Hom{X}{\colim_k Y_k} = \colim_k\Hom{X}{Y_k}$
\end{definition}

\begin{proposition}
  A cocomplete category is small iff it is presentable.
\end{proposition}

\NewDocumentCommand\UU{}{\mathbb{U}}
\NewDocumentCommand\U{}{\mathrm{U}}

\section{Logos}

Assume that $C$ is a category with finite limits (a \emph{lex category}).

\begin{definition}

  A \emph{family} in $C$ parameterized by $B:C$ is a map $\Mor[p]{E}{B}$. A
  morphism of families over the same base is a commuting triangle. The category
  of such things is the slice $C/B$.
%
  The reindexing of a family $\Mor[p]{E}{B}$ along an arbitrary map
  $\Mor{B'}{B}$ is simply the fiber product. From this we construct the functor
  $\Mor[\UU]{\OpCat{C}}{\CAT}$ which takes an object to its slice, and a morphism to
  the fiber product. We call $\UU$ the \emph{universe} of $\CC$.

\end{definition}

\begin{remark}
  In 1-category theory, $\UU$ would be a pseudofunctor into a bicategory.
\end{remark}

\begin{remark}
  The functor $\UU$ gives the fibers of the codomain fibration.
\end{remark}

We define $\U$ as the groupoid core of $\UU$: it has the same objects as $\UU$
but only the invertible arrows. This universe in the sense of type theory; but
neither is actually an object of $C$; they are structures \emph{over} $C$.

Suppose we have a small diagram $\Mor[X]{I}{C}$ and the colimit $X_\infty =
\colim_i X_i$; what if we want to describe a family over the colimit, i.e.\
$\Mor{E}{X_\infty}$? This is the topic of \emph{descent}.


\DeclareFontFamily{U}{min}{}
\DeclareFontShape{U}{min}{m}{n}{<-> dmjhira}{}
\NewDocumentCommand{\hirayo}{}{\text{\usefont{U}{min}{m}{n}\symbol{'110}}}


\NewDocumentCommand\Yo{s g}{%
  \mkern-4mu\hirayo%
  \IfValueT{#2}{%
    \IfBooleanTF{#1}{
      \parens{#2}
    }{#2}
  }%
}

Because groupoids are certain categories, we obtain a ``Yoneda'' embedding
$\Mor[\Yo]{C}{\Hom{\OpCat{C}}{\CAT}}$. The Yoneda lemma says that morphisms
$\Mor[\phi]{\Yo{B}}{\UU}$ are the ``same'' as families $\overline{\phi}\in
\UU(B)=C/B$. $\UU$ classifies families and all morphisms (not only isomorphisms
as $\U$, or only monos as $\Omega$).

Coming back to descent, we can say that families $\Mor{E}{X_\infty}$ are the
same as maps $\Mor{\Yo{X_\infty}}{\UU}$; by reindexing, we obtain maps
$\Mor{X_k}{\UU}$ for each $k:K$. Going the other direction, can we start from
such component families and get a family over the colimit? At first glance
looks like we might be able to, but note that we are constructing a map
$\Mor{\Yo{X_\infty}}{\UU}$ and the representable object on the left only acts
as a colimit with respect to representable objects on the right: but $\UU$ is
not representable.

We want a unique lifting condition: if we define a family out of the
$\Psh{C}$-colimit, can we get a unique map out of the Yoneda embedding of the
$C$-colimit? In other words, can $\UU$ distinguish between the colimit and the
Yoneda embedding of the colimit? This is descent. Phrasing this more precisely, we want
\[
  \Mor{\Hom{\Yo{\colim X}}{\UU}}{\Hom{\colim \Yo{X}}{\UU}}
\]
to be an equivalence of categories. We can pull the colimit out of the hom as a limit, and
then this equivalence is calculated to the more familiar descent condition
$\lim_i\Hom{\Yo{X_i}}{\UU} \simeq \Hom{\Yo{\colim_i{X_i}}}{\UU}$, i.e.\ $\lim_i{(C/X_i)}
\simeq C/{\colim_i X_i}$. There is an obvious adjunction here, given by
reindexing and colimit. \emph{Descent says that this adjunction is an
equivalence.}



\end{document}
